raptr: Representative and Adequate Prioritization Toolkit in R

### 2017-02-13 code/rmarkdown/article.Rmd
update text
address more of the reviewers' concerns

Abstract word count: XXXX / 350  
Total word count: XXXX / 7000  
Number of references: 42  
Number of figures: 8  
Number of tables: 0  
Number of textboxes: 0  
Number of equations: 4  
Short running title: 
Short abstract: Capture biodiversity processes and patterns in prioritizations using raptr https://github.com/jeffreyhanson/raptr  
Keywords: conservation, biodiversity, MILP, reserve-selection, optimization, protected area

@@

1. An underlying aim in conservation planning is to maximize the long-term persistence of biodiversity. To fulfill this aim, the ecological and evolutionary processes that sustain biodiversity must be preserved. One way to conserve such processes at the feature level (eg. species, ecosystem) is to preserve a sample of the feature (eg. individuals, areas) that is representative according to variation in the intrinsic or extrinsic physical attributes that underpin the process of interest. For example, the long-term persistence of a species may depend on its ability to adapt to new climatic conditions. By conserving individuals with traits that are advantageous to their local environment--physical attributes associated with adaptation--protected areas can foster the process of evolutionary adaptation. Despite this, current reserve selection methods overwhelmingly focus on securing an adequate amount of individuals or habitat for each feature. Little attention has been directed towards capturing a representative sample of each feature.

2. To address this issue, we developed the \texttt{raptr R} package to guide reserve selection using targets for representing spatially explicit variables that underpin biodiversity processes as well as more typical area protection targets. Users set "amount targets"--similar to conventional methods--to secure a sufficient proportion of the features' geographic distributions. Additionally, users set "space targets" to secure a representative sample of variation in ecologically or evolutionary relevant attributes (eg. climatic or genetic variation). We demonstrate the functionality of this package using simulations and two case studies. We generate solutions using just amount targets--representing conventional methods--and compare them with solutions generated using amount and space targets.

3. We show that different solutions emerge when explicitly considering within feature representation in reserve selection. Our simulations suggest that targeting representativeness is particularly important for biodiversity features that have multimodal distributions in an attribute space. The case studies show that setting space targets can result in solutions that conserve a much more representative sample of features--and this can be achieved with only a slight increase in reserve size.

4. The \texttt{raptr R} package provides a toolkit for making spatial conservation prioritizations that secure an adequate and representative sample of features. By targeting within-feature representativeness, prioritizations may have a greater chance of achieving long-term biodiversity persistence.

>>>>

Abstract word count: 347 / 350  
Total word count: XXXX / 7000  
Number of references: 42  
Number of figures: 8  
Number of tables: 0  
Number of textboxes: 0  
Number of equations: 4  
Short running title: 
Short abstract: Prioritize biodiversity processes and patterns using raptr https://github.com/jeffreyhanson/raptr  
Keywords: conservation, biodiversity, mixed integer linear programming, optimization, protected areas

@@

1. An underlying aim in conservation planning is to maximize the long-term persistence of biodiversity. To fulfill this aim, the ecological and evolutionary processes that sustain biodiversity must be preserved. One way to conserve such processes at the feature level (eg. species, ecosystem) is to preserve a sample of the feature (eg. individuals, areas) that is representative in terms of the intrinsic or extrinsic physical attributes that underpin the process of interest. For example, by conserving a sample of populations with local adaptations that is representative in terms of the full variation found in the species--physical attributes associated with adaptation--protected areas can foster adaptative processes by ensuring these adaptations are not lost. Despite this, current reserve selection methods overwhelmingly focus on securing an adequate amount of area or habitat for each feature. Little attention has been directed towards capturing a representative sample of each feature.

2. To address this issue, we developed the \texttt{raptr R} package to help guide reserve selection. Users set "amount targets"--similar to conventional methods--to ensure that solutions secure a sufficient proportion of area or habitat for each feature. Additionally, users set "space targets" to secure a representative sample of variation in ecologically or evolutionary relevant attributes (eg. environmental or genetic variation). We demonstrate the functionality of this package using simulations and two case studies. In these studies, we generated solutions using amount targets--representing conventional methods--and compared them with solutions generated using amount and space targets.

3. Our results demonstrate that different solutions emerge when targeting a representative sample of each feature. We show that using these targets is important for features that have multimodal distributions in the process-related attributes (eg. species with multimodal niches). We also found that solutions could conserve a far more representative sample with only a slight increase in reserve size.

4. The \texttt{raptr R} package provides a toolkit for making prioritizations that secure an adequate and representative sample of each feature. By using solutions that secure a representative sample of each feature, prioritizations may have a greater chance of achieving long-term biodiversity persistence.

### 2017-02-13 code/rmarkdown/text.Rmd
address more of the reviewers' concerns

## Optimization
The unreliable formulation is non-linear. However, the non-linear components can be linearized using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearized using methods described by Beyer et al. [-@r426]. The \texttt{raptr} R package provides functions to express conservation planning data as optimization problems using linearized versions of the unreliable and reliable formulations (see Appendix S1 for reliable formulation), and solve them using the commercial \texttt{Gurobi} software suite (\url{www.gurobi.com}). Presently, academics can obtain a license at no cost from the \texttt{Gurobi} website.

@@

To understand and showcase the behavior of the unreliable problem, we conducted a simulation study and two case studies. These studies involved generating solutions using only amount targets--equivalent to solutions generated using conventional methods (eg. \texttt{Marxan})--and solutions using amount and space targets using the unreliable formulation. By comparing these solutions, we can gauge the benefits of explicitly including space targets in reserve selection. We have also conducted a benchmark analysis to estimate how long it would take to solve various sized problems (Appendix S2). All analyses were performed in \texttt{R} [version `r getRversion()`; @r192] and all optimization problems were solved to within `r general.params.LST[[MODE]]$MIPGap * 100` % of optimality.

## SIMULATION STUDY
### Methods
We simulated a hypothetical study area with planning units arranged in a $`r sqrt(sim.params.LST[[MODE]]$number.planning.units)` \times `r sqrt(sim.params.LST[[MODE]]$number.planning.units)`$ square grid (Fig. 2). We then simulated three species across the study area. The first species was simulated to represent a hyper-generalist--occurring in all planning units with equal probability (Fig. 2a). This species' distribution was based on a uniform distribution (eqn 4a). The second species was simulated to represent a species with an idealized distribution (Fig. 2b). It has a core range area, and is less likely to be found in areas that are distant from the core area. This species' distribution was simulated using the density function of a single multivariate normal distribution (represented by $\mathcal{N}$; eqn 4b). The third species was simulated to represent a species with two distinct populations (Fig. 2c). It was simulated using the maximum of two density of two multivariate normal distributions (eqn 4c). The probability that each species inhabited a given planning unit was calculated using the $X, Y$ coordinates of the planning unit and eqns 4a--4c. We used a geographic attribute space to showcase the behavior of the problem. For each species, demand points were set as the planing units' centroids and were weighted according to the units' probability of occupancy.

@@

We generated four solutions for each species. First to represent solutions generated using conventional methods, we generated solutions using only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. Second to show how the addition of space-based targets can affect a prioritization, we generated solutions using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % space targets to ensure that the solution samples the geographic spread of the species (hereafter referred to as geographic spread targets). Third to represent solutions generated using conventional planning methods that penalize for fragmentation, we generated solutions using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`. Fourth to illustrate the combined affect using space-based targets and penalising fragmentation, we generated solutions using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic spread targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`.

@@

The uniform species occurred in all planning units with equality probability (Fig. 2a). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount-based targets (Fig. 3a) selected `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area. This specific configuration is an artifact of the method used to solve the problem. In fact, for this species, all prioritizations containing `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units are optimal when considering only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. In the absence of additional criteria, reserve selection methods may or may not return solutions that secure a representative sample of the features. In this particular case, the solution does not secure a representative sample of the species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). Note that the proportion here is negative because it performs worse than simply siting a planning unit in the centroid of the species' demand points (equivalent to the $R^2$ of a regression model that explains less variation in the response variable than the mean of the response variable).

The addition of a `r sim.params.LST[[MODE]]$space.target*100` % geographic spread target resulted in a solution that secured a representative sample of the uniform species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % sampled; Fig. 3b). Since all planning units have equal cost and an equal chance of being occupied, this solution has the same number of planning units as the solution generated using only amount targets (cf. Fig 3a). Although the use of amount and space targets has yielded a solution that conserves a sample of the species that contains an adequate amount of habitat and a representative sample of the species' geographic spread, they have yielded a solution that is also highly fragmented solution.

The addition of a positive boundary length modifier (BLM) parameter resulted in a well-connected solution that secured an adequate proportion (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount &\nspace targets')$amount.held,2)` % ) and representative sample (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount & space\ntargets and BLM')$space.held,2)` % sampled) of the species geographic spread (Fig. 3d). Although this solution also contains `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount & space\ntargets and BLM')$n` planning units--the same amount as previously discussed solutions--it secures a representative proportion of the species geographic spread in a configuration that is not highly fragmented. Overall, these results show that under the simplest of conditions, using just amount targets results in an under-specified reserve selection problem that is unlikely to return an effective solution.

@@

The normally distributed species was simulated to contain a single core area where individuals were most prevalent, and marginal areas where individuals were less likely to occur (Fig. 2b). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount targets contained `r filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount\ntargets')$n` planning units and concentrated conservation efforts in the core area (Fig. 3e). Since all planning units have equal costs and areas, this solution contains the planning units with the highest probabilities of occupancy. Whilst this solution conserves a sufficient amount of habitat in a cost-effective manner--it fails to secure a representative sample of the species' geographic spread (`r round(filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % distribution sampled).

The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount targets and `r sim.params.LST[[MODE]]$space.target*100` % geographic spread targets resulted in a solution that is both adequate and representative in terms of the uniform species' distribution (Fig. 3f). This solution used `r filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount &\nspace targets')$n` planning units to secure `r round(filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount &\nspace targets')$amount.held,2)` % and sample `r round(filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % of the normal species' distribution. By using amount and geographic spread targets in the reserve selection problem, we obtained a solution that concentrates conservation effort in the range core and also the range margin. However, similar to the corresponding solution for the uniform species (Fig. 3b), this solution is also highly fragmented.

By using `r sim.params.LST[[MODE]]$amount.target*100` % amount targets and `r sim.params.LST[[MODE]]$space.target*100` % geographic spread targets and a boundary length modifier parameter ($BLM = 1$), we obtained a solution that fulfills adequacy, representativeness and connectivity objectives (Fig. 3h). This solution contains `r filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount & space\ntargets and BLM')$n` planning units. Unlike the solution for the uniform species generated using the same parameters, however, this solution contains more planning units than any other solutions for this species (cf. Figs. 3e--3g). These results suggests that the combination of space targets and boundary length modifiers may require solutions that contain more planning units. They also also suggest that space targets can result in solutions that secure a more representative sample of the species--even for species without significant spatial structure.

@@

The bimodally distributed species was simulated to represent a species with a highly structured distribution, and it contains two distinct populations (Fig. 2c). The solution generated using just `r sim.params.LST[[MODE]]$amount.target*100` % amount targets prioritized individuals in just one of the two populations (Fig. 3i). While this solution secured an adequate proportion of the species' distribution (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount\ntargets')$amount.held,2)` % secured), it did not sample a representative proportion of the range (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). The addition of boundary length modifiers to amount targets resulted in a solution which sampled even less of the species distribution (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount target\nand BLM')$space.held,2)` % sampled; Fig. 3k). However, the addition of geographic spread targets resolved these issues.

The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount and `r sim.params.LST[[MODE]]$space.target*100` % geographic spread targets included planning units from both populations (Fig. 3j). Although this solution required more planning units to fulfill both targets ($n=`r filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount &\nspace targets')$n`$), this solution secured a representative sample of the bimodally distributed species (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % range sampled). Similar to corresponding solution generated for the normally distributed species (Fig. 3f), this solution was also highly fragmented and we could obtain a more well-connected solution at the expense of selecting additional planning units ($n=`r filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount & space\ntargets and BLM')$n`$; Fig. 3l). The results for the bimodally distributed species suggest that species with highly structured populations or strong variation between individuals could benefit the most from prioritizations that are generated using space-based targets.

@@

# Implications and future directions
The \texttt{raptr R} package provides a unified approach to reserve selection. Conservation planners can use this \texttt{R} package to generate prioritizations that secure intra-specific as well as as inter-specific biodiversity patterns. Additionally, the package contains functionality to accommodate uncertainty in the distribution of features using the reliable formulation (defined in Appendix S1), and ensure that solutions are well-connected. Both the simulated and case study species suggest that explicitly considering representation as well as amount targets can significantly alter prioritizations.

@@

Attribute data need to be spatially comprehensive to map planning units to attribute spaces. However, many real-world data sets are patchy. For example, due to constrained resources, genetic data may only be available for some planning units. To use such patchy data, planners could omit the planning units that are missing data, or estimate the missing data using spatially explicit models [eg. krigging or generalized dissimilarity models; @r467; @r486]. This approach has been successfully applied to a range of biological data sets [eg. @r468]. We caution that inaccuracy and uncertainty in data can negatively impact prioritizations [@r488], and recommend that planners ensure that data are reliable or accommodate uncertainty into the planning process (eg. by using the reliable formulation).

To maximize the long-term persistence of biodiversity, decision makers need to identify prioritizations that preserve existing patterns of biodiversity and the processes that support them.  To achieve this, conservation planners need a decision support tool that can deliver cost-effective prioritizations that secure an adequate amount of a representative sample of biodiversity features. Here, we developed the \texttt{raptr R} package to fill this void. By exploring the functionality of this package using several simulated and case study species, we found that including space-based targets can result in substantially different solutions.

>>>>

## Optimization
Although the reserve selection problems presented here are non-linear (see Appendix S1 for reliable formulation), they can be linearized using methods described by Beyer _et al._ [-@r426] and Cui _et al._ [@r16]. The \texttt{raptr} R package provides functions to express conservation planning data as linearized versions of the optimization problems and solve them using the commercial \texttt{Gurobi} software suite (\url{www.gurobi.com}). Presently, academics can obtain a license at no cost.

@@

To showcase the behavior of the unreliable formulation, we conducted a simulation study and two case studies. To understand how long it would take to solve various sized problems, we also conducted a benchmark analysis (Appendix S2). We completed the analyses using \texttt{R} [version `r getRversion()`; @r192] and solved all optimization problems to within `r general.params.LST[[MODE]]$MIPGap * 100` % of optimality.

## SIMULATION STUDY
### Methods
We simulated a hypothetical study area with square planning units arranged in a $`r sqrt(sim.params.LST[[MODE]]$number.planning.units)` \times `r sqrt(sim.params.LST[[MODE]]$number.planning.units)`$ grid (Fig. 2). We then simulated three species to inhabit this study area. Firstly, we simulated a hyper-generalist species (hereafter referred to as the "uniformly distributed species"). It all planning units with equal probability (Fig. 2a; eqn 4a). Secondly, we simulated a species with simple habitat requirements (hereafter referred to as the "normally distributed species"; Fig. 2b). This species was most likely to be found in planning units in the centre of the study area, and was simulated using density function of a multivariate normal distribution (represented by $\mathcal{N}$; eqn 4b). Thirdly, we simulated a species two distinct populations (hereafter referred to as the "bimodally distributed species"; Fig. 2c). It was simulated using the maximum of two density of two multivariate normal distributions (eqn 4c). Planning unit occupancy was calculated using the $(X, Y)$ coordinates of the units' centroids and the corresponding equations. We used a geographic attribute space to provide an intuitive visualization of the solutions. Demand points were set as the planning units' centroids and were weighted according to the units' probability of occupancy.

@@

We generated four solutions for each species. Firstly, to represent solutions generated using conventional methods (eg. \texttt{Marxan}), we generated solutions using `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. Secondly, to show how the addition of space targets can affect solutions, we generated solutions using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % space targets to capture the geographic spread of each species (hereafter referred to as geographic spread targets). Thirdly, to represent solutions generated using conventional planning methods that penalize for fragmentation, we generated solutions using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`. Fourthly, to illustrate the combined effects of using amount and space targets as well as penalties for fragmentation, we generated solutions using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic spread targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`.

@@

The solution generated for the uniformly distributed species using an amount target prioritized `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area (Fig. 2a). For this species, all solutions containing this many planning units are optimal when using a `r sim.params.LST[[MODE]]$amount.target*100` % amount target to constrain the reserve selection problem. We obtained this particular solution due to artifacts in the solver. This solution captured a small amount of the geographic spread of the species (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). In fact, the coverage was so poor that the proportion was negative because the solution captured a less representative sample than a solution containing one planning unit in the centre of the species' distribution.

By explicitly targeting representativeness, we obtained a solution that captured the geographic spread of the uniform species (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % sampled; Fig. 3b). Although this solution secured an adequate amount of habitat and a representative sample of the species' geographic spread, this solution was highly fragmented. However, by penalizing fragmentation using a boundary length modifier, we were able to obtain a well-connected solution that met all of the objectives (Fig. 3d). This solution prioritized the same number of planning units as the previous solutions even though it is far superior. As we can see, under the simplest of circumstances, reserve selection methods may not yield solutions that secure a representative sample of features unless constraints are used to guarantee this objective.

@@

The solution generated for the normally distributed species using just an amount target prioritized planning units in the species' core range area (Fig. 3e). While this strategy is cost-effective for protecting habitat, it was ineffective for securing a representative sample of the species' geographic spread (`r round(filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled) because it did not protect any planning units along the periphery of the species' distribution. By using amount and space targets, we were able to conserve an adequate amount of habitat and also secure a representative sample of the species' geographic spread (`r round(filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % sampled; Fig. 3f). Similar to the solutions for the uniformly distributed species, this solution was highly fragmented and we were able to obtain a better connected solution by specifying fragmentation penalties (Fig. 3h). However, unlike the solutions for the uniformly distributed species, the solution generated using an amount target, a space target, and fragmentation penalties required more planning units than the other solutions. These results suggest that solutions may need to prioritize more planning units to meet additional conservation objectives.

@@

The solution generated for the bimodally distributed species using an amount target only conserved individuals belonging to one of the two populations (Fig. 3i). As a consequence, this solution did not secure a representative sample of the species' geographic spread (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). The addition of fragmentation penalties exacerbated this issue, and resulted in a solution that sampled even less of the species' geographic spread (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount target\nand BLM')$space.held,2)` % sampled; Fig. 3k). However, this issue was resolved by using a space target  to obtain a solution that secured both populations (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount & space\ntargets and BLM')$space.held,2)` % geographic spread secured; Fig. 3l). This finding suggests that species with large intra-specific variation could benefit the most from prioritizations generated using space-based targets.

@@

# Implications and future directions
The \texttt{raptr R} package provides a unified approach to reserve selection. Conservation planners can use this \texttt{R} package to generate prioritizations that secure intra- and inter-specific biodiversity patterns. Both the simulations and case studies show that explicitly targeting a representative sample of each feature in a conservation planning exercise can substantially alter prioritizations. Additionally, we found that prioritizations often needed to secure more habitat to conserve a representative sample of each feature. Although it may not be practical to conserve substantially more habitat than the amount specified using amount-based targets [eg. 10 % geographic range; @r255], we did find that even small increases in reserve size could yield large gains.

@@

Attribute data need to be spatially comprehensive to map planning units to attribute spaces. However, many real-world data sets are patchy. For example, due to constrained resources, genetic data may only be available for some planning units. To use such patchy data, planners could omit the planning units that are missing data, or estimate the missing data using spatially explicit models [eg. krigging or generalized dissimilarity models; @r467; @r486]. This approach has been successfully applied to a range of biological data sets [eg. @r468]. We caution that inaccuracy and uncertainty in data can negatively impact prioritizations [@r488], and recommend that planners ensure that data are sufficiently reliable or use the reliable formulation (described in Appendix S1) to accommodate uncertainty into the reserve selection process.

To maximize the long-term persistence of biodiversity, decision makers need to identify prioritizations that preserve existing patterns of biodiversity and the processes that support them. It is becoming increasingly apparent that simply protecting a large area will fail to achieve this [@r492]. Here, we developed the \texttt{raptr R} package to provide conservation planners with the tools to deliver cost-effective prioritizations that secure an adequate amount of a representative sample of biodiversity features. By exploring the functionality of this software package, we found that explicitly targeting a representative sample of each feature can result in substantially different solutions.
